---
title: "Getting Started with GeneCardAnalysis"
author: "Sebastiano Di Bella, Francesco Orilio"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Getting Started with GeneCardAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

## Overview

**GeneCardAnalysis** is an R Shiny application for reproducible analysis and
visualisation of RT-qPCR data. It implements the delta-CT normalisation method
(Livak & Schmittgen, 2001) to compute:

- **Normalised expression**: 2^(−ΔCT), relative to one or more housekeeping genes
- **Log2 fold-change (log2FC)**: log2(sample / reference condition)
- **Z-score**: per-gene standardisation on the log2 scale (= −ΔCT), suitable
  for comparing relative expression patterns across many samples

Results are exported as publication-ready heatmaps (PNG, PDF) and Excel tables.

This vignette demonstrates the core computational functions that underlie the
app, using a minimal toy dataset with known values so the reader can verify the
maths by hand.

---

## Installation

```{r install, eval = FALSE}
# CRAN dependencies
install.packages(c(
  "shiny", "bslib", "circlize", "readxl", "tidyverse",
  "grid", "glue", "writexl", "showtext", "yaml",
  "DT", "colourpicker", "gtools"
))

# Bioconductor dependency
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")

# Install GeneCardAnalysis from GitHub
# install.packages("remotes")
remotes::install_github("StassiLab/GeneCardAnalysis")
```

```{r load-pkg}
# Source directly if running from the package root
source("../card_functions.R")
source("../signatures_genes_card.R")

library(dplyr)
library(ggplot2)
```

---

## 1. Input data format

The app accepts a CSV (or TSV) file with at least three columns:

| Column        | Description                                           |
|---------------|-------------------------------------------------------|
| `Sample.Name` | Unique sample identifier (e.g. `UT`, `EMT_24h`)      |
| `Target.Name` | Gene symbol (e.g. `VIM`, `CDH1`, `GAPDH`)            |
| `CT`          | Cycle-threshold value, numeric or `"Undetermined"`    |

This format is the default export of Applied Biosystems QuantStudio and
Bio-Rad CFX Manager. Both comma (`,`) and semicolon (`;`) separators are
auto-detected by the app.

### Toy dataset

We build a minimal dataset — 2 samples, 4 genes (including housekeeping) —
with CT values chosen so the output can be verified analytically:

```{r toy-data}
ct_data <- data.frame(
  Sample.Name = rep(c("UT", "EMT"), each = 4),
  Target.Name = rep(c("GAPDH", "VIM", "CDH1", "SNAI1"), times = 2),
  CT = c(
    20, 25, 30, 27,   # UT:  GAPDH=20, VIM=25, CDH1=30, SNAI1=27
    20, 23, 32, 24    # EMT: GAPDH=20, VIM=23, CDH1=32, SNAI1=24
  ),
  stringsAsFactors = FALSE
)

ct_data
```

---

## 2. Delta-CT normalisation

`extract.normalised.CTs()` applies the delta-CT method and returns a
**genes × samples** matrix of 2^(−ΔCT) values.

```{r normalise}
norm_mat <- extract.normalised.CTs(
  CTs               = ct_data,
  housekeeping.gene = "GAPDH",   # explicit; auto-selects GAPDH/HPRT1 if NULL
  undetermined      = 40,        # value for "Undetermined" calls
  handle.NAs        = TRUE       # impute NAs near the observed minimum
)

round(norm_mat, 6)
```

### Verification

For `VIM` in the `UT` sample:

$$
\Delta CT = CT_{VIM} - CT_{GAPDH} = 25 - 20 = 5
$$
$$
2^{-\Delta CT} = 2^{-5} = 0.03125
$$

```{r verify}
expected_VIM_UT <- 2^(-(25 - 20))
cat("Expected VIM/UT:", expected_VIM_UT, "\n")
cat("Computed VIM/UT:", norm_mat["VIM", "UT"], "\n")
cat("Match:", isTRUE(all.equal(norm_mat["VIM", "UT"], expected_VIM_UT)), "\n")
```

Note that `GAPDH` is absent from the output — it is used for normalisation
and then removed.

---

## 3. Log2 fold-change

Log2FC is computed as log2(sample / reference), where `UT` is the reference:

```{r log2fc}
ref_col   <- "UT"
log2fc_mat <- log2(norm_mat / norm_mat[, ref_col])
log2fc_mat[is.infinite(log2fc_mat)] <- NA

round(log2fc_mat, 4)
```

Interpretation:

- `VIM` log2FC = `r round(log2fc_mat["VIM", "EMT"], 3)` in EMT vs UT →
  VIM is **upregulated** (~`r round(2^log2fc_mat["VIM", "EMT"], 1)`× more expressed)
- `CDH1` log2FC = `r round(log2fc_mat["CDH1", "EMT"], 3)` in EMT vs UT →
  CDH1 is **downregulated** (consistent with EMT)

---

## 4. Z-score (log2 scale)

> **Important**: z-scores must be computed on the **log2 scale**
> (i.e. log2(2^(−ΔCT)) = −ΔCT), not on the raw 2^(−ΔCT) values.
> Computing z-scores on exponential data violates the normality assumption
> of the standardisation and distorts patterns towards highly-expressed genes.

```{r zscore}
mat_log2 <- log2(norm_mat)
mat_log2[!is.finite(mat_log2)] <- NA

# Per-gene z-score across samples (scale rows)
zmat <- t(scale(t(mat_log2)))
round(zmat, 3)
```

---

## 5. Gene signatures

The package ships with four built-in gene panels:

```{r signatures}
sigs <- get_signatures()
names(sigs)

# Number of genes in each panel
sapply(sigs[sigs != "all"], length)
```

Custom signatures can be defined in the YAML configuration file:

```yaml
custom_signatures:
  MyEMTPanel: ["VIM", "CDH1", "SNAI1", "ZEB1", "TWIST1"]
```

And accessed programmatically:

```{r custom-sig}
cfg <- list(
  custom_signatures = list(
    MyEMTPanel = c("VIM", "CDH1", "SNAI1", "ZEB1", "TWIST1")
  )
)
sigs_custom <- get_signatures(cfg)
sigs_custom$custom_MyEMTPanel
```

`build_signature_choices()` returns a named vector ready for use in
`shiny::selectInput()`:

```{r sig-choices}
build_signature_choices(cfg)
```

---

## 6. Raw CT extraction

`extract.CTs()` returns the raw CT matrix (no normalisation), useful for
quality control and visual inspection of housekeeping gene stability:

```{r raw-cts}
raw_mat <- extract.CTs(ct_data)
round(raw_mat, 2)
```

### Housekeeping gene stability

A simple CV (coefficient of variation) check across samples:

```{r hk-stability}
hk_cts <- raw_mat["GAPDH", ]
cat(sprintf(
  "GAPDH — mean CT: %.2f, SD: %.3f, CV: %.1f%%\n",
  mean(hk_cts), sd(hk_cts), 100 * sd(hk_cts) / mean(hk_cts)
))
```

A CV < 1% indicates excellent housekeeping gene stability, as is the case
here (GAPDH is identical across both samples in this toy example).

---

## 7. Running the Shiny app

```{r launch-app, eval = FALSE}
shiny::runApp(".")          # from the GeneCardAnalysis directory
# or, once installed as a package:
GeneCardAnalysis::run_app()
```

The app accepts a YAML configuration file that defines all experimental
comparisons. A minimal example:

```yaml
project_name: EMT_study
author: SDB
global_settings:
  housekeeping_genes: [GAPDH]
  undetermined_value: 40
  exclude_genes: [RT, gDNA, GDNA, PCR]
comparisons:
  - name: EMT_vs_UT
    reference: UT
    samples: [UT, EMT]
    signature_type: signature_emt
    cluster_heatmap: true
```

---

## Session information

```{r session-info}
sessionInfo()
```

---

## References

Livak KJ, Schmittgen TD (2001). "Analysis of relative gene expression data
using real-time quantitative PCR and the 2^(−ΔΔCT) Method."
*Methods*, **25**(4), 402–408. doi:[10.1006/meth.2001.1262](https://doi.org/10.1006/meth.2001.1262)

Gu Z, Eils R, Schlesner M (2016). "Complex heatmaps reveal patterns and
correlations in multidimensional genomic data."
*Bioinformatics*, **32**(18), 2847–2849. doi:[10.1093/bioinformatics/btw313](https://doi.org/10.1093/bioinformatics/btw313)
